---
output:
  html_document:
    css: ./knitr.css
    fig_caption: yes
    highlight: default
    keep_md: no
    number_sections: no
    theme: journal
    toc: yes
---

May 2015 - master -- trainr

```{r, message=FALSE}
library('trainr')
path <- system.file(package = 'trainr')
```

```{r frontmatter, child="./01-frontmatter.Rmd", include=FALSE}
```

```{r data, include=FALSE}
opts_chunk$set(echo = TRUE)

# ## read in all csv files in dir
# lf <- list.files(file.path(path, 'extdata', 'csv'), full.names = TRUE,
#                  pattern = '\\.csv')
# l_dat <- setNames(lapply(lf, read.csv, stringsAsFactors = FALSE,
#                          strip.white = TRUE, header = TRUE),
#                   gsub('\\.csv', '', basename(lf)))


## or generate data based on number of pts
N <- 35
source(file.path(path, 'scripts', '00-gen_data.R'))


merge_all <- function(l, ...)
  ## recursively merge list of data frames
  ## usage: merge_all(list(dd1, dd2), by = 'id')
  ## params:
    ## l: a list of data frames
    ## ... additional arguments passed to ?merge
  Reduce(function(x, y) merge(x, y, ...), l)
f_dt <- function(x) as.Date(x, origin = '1970-01-01')

dd <- merge_all(list(demo, resp, surv, fwup), by = 'id', all = TRUE)

dd <- dd1 <- within(dd, {
  ## format dates
  dt_dob <- dmy(d_dob, m_dob, y_dob)
  dt_reg <- dmy(d_reg, m_reg, y_reg)
  dt_resp <- dmy(d_resp, m_resp, y_resp)
  dt_status <- dmy(d_status, m_status, y_status)
  dt_fwup <- dmy(d_fwup, m_fwup, y_fwup)
  dt_lastcontact <- f_dt(ifelse('Dead' %in% status, dt_status, dt_fwup))
  
  
  ## calculate some useful stuff
  stage <- as.character(as.roman(stage))
  age_dx <- floor(as.numeric(dt_reg - dt_dob) / 365.242)
  age_cat <- cut(age_dx, breaks = c(0, 50, 60, 70, 80, 100),
                 labels = c(sprintf('%s - 50', min(age_dx)),
                            '51 - 60', '61 - 70', '71 - 80',
                            sprintf('81 - %s', max(age_dx))))
  
  
  ## ipi factors: age > 60, ps > 1, elevated ldh, ens > 1, stage 3/4
  age_ipi <- c('&le; 60','&gt; 60')[(age_dx > 60) + 1L]
  ecog_ipi <- c('&le; 1','&gt; 1')[(ecog > 1) + 1L]
  ldh_ipi <- ldh
  ens_ipi <- c('&le; 1','&gt; 1')[(ens > 1) + 1L]
  stage_ipi <- c('Stage I/II','Stage III/IV')[(stage %in% c('III', 'IV')) + 1L]
  
  ipi <- psum(age_dx > 60, ecog > 1, ldh %in% 'Elevated',
              ens > 1, stage %in% c(3, 4))
  ipi_risk <- c('Low','Low','Low-Int','High-Int','High','High')[ipi + 1]
  ipi_risk <- factor(ipi_risk, levels = c('Low','Low-Int','High-Int','High'))
  
  
  ## survival
  os_ind <- ('Dead' %in% status) + 0L
  dt_os <- f_dt(ifelse('Dead' %in% status, dt_status, dt_lastcontact))
  os_time <- as.numeric(dt_os - dt_reg) / 30.4
  
  pfs_ind <- (('PD' %in% resp) + 0L) | (os_ind %in% 1)
  dt_pfs <- f_dt(ifelse('PD' %in% resp, dt_resp, dt_os))
  pfs_time <- as.numeric(dt_pfs - dt_reg) / 30.4
})


## table/figure counters:
options(figure_counter = TRUE,
        figure_counter_roman = TRUE)


## prints n and percent of total N
nnr <- function(x) sprintf('%s (%s%% of)', num2char(x), round(x / N * 100))
```

## Administrative Information

From `r paste0(range(format(dd$dt_reg, '%B %Y')), collapse = ' to ')`, `r N` patients were enrolled on study at `r print_counts(dd$site)`. As per inclusion criteria, all patients were newly diagnosed diffuse large B-cell lymphoma (DLBCL). By study design patients were stratified by subtype, and all results were performed as such.

`r nnr(N)` patients received at least one dose of study treatment and were included in toxicity and response analyses.

## Patient Characteristics

Patient characteristics are presented in Table 1. `r nnr(sum(dd$sex %in% 'Male'))` patients enrolled were male. The median age at diagnosis was `r intr(dd$age_dx)`. `r nnr(sum(dd$ipi > 1))` patients had International Prognostic Index (IPI) of two or more.

```{r, results='asis'}
by_var <- 'subtype'
dd1[] <- lapply(dd1, function(x) if (is.character(x)) factor(x) else x)

## wrapper functions to make tables
median_minmax <- function(...) describeMedian(..., iqr = FALSE)

get_stat <- function(var, by = by_var, data = dd1, digits = 0) {
  getDescriptionStatsBy(data[, var], by = data[, by], percentage_sign = '',
                        useNA = 'ifany', add_total_col = TRUE,
                        show_all_values = TRUE, hrzl_prop = FALSE,
                        statistics = FALSE, html = TRUE, digits = digits,
                        continuous_fn = median_minmax)
}

make_table <- function(x, ...)
  htmlTable(x, ..., ctable = TRUE, align = 'ccc', title = '',
            cgroup = c('', 'Subtype'), n.cgroup = c(1, 2),
            css.tspanner = "font-weight: 900; text-align: center;")

## table headers common to (most) tables
tbl_headers <- c(sprintf('All patients<br /><font size=1>N = %s</font>', N),
                 sprintf('%s<br><font weight=normal; size=1>n = %s</font>',
                         levels(dd1[[by_var]])[1],
                         sum(dd1[[by_var]] == levels(dd1[[by_var]])[1])),
                 sprintf('%s<br><font weight=normal; size=1>n = %s</font>',
                         levels(dd1[[by_var]])[2],
                         sum(dd1[[by_var]] == levels(dd1[[by_var]])[2])))


## select variables to use in table 1
tbl <- list()
## clin vars
tbl[['Site']] <- get_stat('site')
tbl[['Sex']] <- get_stat('sex')
tbl[['Race']] <- get_stat('race')
tbl[['Ethnicity']] <- get_stat('ethnic')
## ipi 
tbl[['Age at diagnosis']] <- rbind(get_stat('age_dx'),
                                   get_stat('age_cat'))
tbl[['ECOG PS']] <- rbind(get_stat('ecog'),
                          get_stat('ecog_ipi'))
# tbl[['Disease stage']] <- rbind(get_stat('stage'),
#                                 get_stat('stage_ipi'))
tbl[['Disease stage']] <- get_stat('stage_ipi')
tbl[['LDH']] <- get_stat('ldh')
tbl[['ENS']] <- rbind(get_stat('ens'),
                      get_stat('ens_ipi'))
tbl[['IPI']] <- get_stat('ipi_risk')


## combine summaries into matrix and add total row at bottom
out <- rbind(do.call(rbind, tbl),
             c(nrow(dd), sprintf('%s (%s)', table(dd$subtype),
                                 round(table(dd$subtype) / N * 100))))
## some formatting for table
rgroup <- names(tbl)
n.rgroup <- unname(sapply(rgroup, function(x) nrow(tbl[[x]])))
n.tspanner <- sapply(list(1:4, 5:10),
                     function(x) nrow(do.call('rbind', tbl[x])))

colnames(out) <- tbl_headers
## 0 counts/percents make table harder to read
out <- gsub('0 (0)', '-', out, fixed = TRUE)


make_table(out, rgroup = c(rgroup, 'Total'), n.rgroup = c(n.rgroup, 1),
          tspanner = c('','IPI factors',''), n.tspanner = c(n.tspanner, 1),
          tfoot = paste0('<font size=1><sup>&dagger;</sup>Some note.<br />',
                         'Abbreviations: IPI, international prognostic score; ',
                         'ECOG PS, Eastern Cooperative<br />Oncology Group ',
                         'performance score.</font>'),
          caption = 'Table 1: Patient characteristics.')
```

## Baseline labs

Lab results at study entry are presented in Table 2. The median sum of largest tumor diameters at baseline was `r median(blabs$ltd)` for all patients and `r paste0(round(with(blabs, tapply(ltd, dd$subtype, median)), 1), collapse = ' and ')` for ABC and GCB suptype patients, respectively.

```{r, results='asis'}
blabs1 <- merge(blabs, dd[, c('id','subtype')])
blabs1$id <- NULL
nums <- sapply(blabs1, is.numeric)
nums <- names(nums[nums])

## calculate statistics only for the numeric lab variables
out <- setNames(lapply(names(blabs1[, nums]), function(x)
  get_stat(x, data = blabs1, digits = 1)), nums)
out <- do.call('rbind', out)
colnames(out) <- tbl_headers

## formatting
nums <- ifelse(nchar(nums) < 4, toupper(nums),
               gsub('(.)(.*)', '\\U\\1\\L\\2', nums, perl = TRUE))


make_table(out, rgroup = nums, n.rgroup = rep(1, length(nums)),
           caption = 'Table 2: Baseline lab results')
```

```{r, fig.cap='Percent change in hemoglobin values from baseline to end of treatment. ABC and GCB patients shown in grey and blue, respectively.'}
hgb_change <- within(merge(blabs, demo[, c('id','subtype')]), {
  change <- (hgb - elabs$hgb) / hgb
})
hgb_change <- hgb_change[order(hgb_change$change), ]


## waterfall plot
par(mar = c(3, 5.5, 2, .5), family = 'HersheySerif',
    las = 1, tcl = .2)
with(hgb_change,
     barplot(change, border = NA, xlab = 'Patients', 
             ylab = 'Change in HgB (%)\nfrom baseline',
             col = ifelse(subtype %in% 'ABC', 'grey50', 'dodgerblue2')))
mtext('Patients', side = 1)
box('outer')
```

## Treatment toxicities

```{r}
## get tox code info
tox <- cbind(tox, rawr::match_ctc(tox$tox_code)$matches[, c('tox_desc','tox_cat')])


# ## get worst grades by id, toxicity code
# tox1 <- tox[with(tox, order(id, tox_desc, -xtfrm(tox_grade))), ]
# tox1 <- tox1[with(tox1, !duplicated(cbind(id, tox_desc))), ]
# 
# tox2 <- tox_worst(tox, 'id')$tox_worst
# identical(tox1, tox2)


worst <- tox_worst(tox)$tox_worst
worst <- merge(worst, dd[, c('id','subtype')])
worst <- within(worst, {
  tox_desc <- factor(tox_desc)
  subtype <- factor(subtype)
  tox_grade12 <- ifelse(tox_grade %in% 1:2, 0, tox_grade)
  tox_grade12 <- factor(tox_grade12, levels = c(0, levels(tox_grade)),
                        labels = c('1 - 2', levels(tox_grade)))
  ## comment out to separate levels 1,2
  ## repeat similar to combine/separate levels 3,4 or others
  tox_grade <- droplevels(tox_grade12)
  worst_overall <- ave(as.numeric(tox_grade), id, FUN = max)
})


n <- table(dd$subtype)
## worst tox grade overall per pt
nn <- table(with(worst, tapply(as.numeric(tox_grade), id, max)))


tt <- list(tabler_by(worst, 'tox_desc', 'subtype', n,
                     zeros = '-')[, 1, drop = FALSE],
           tabler_by(worst[worst$subtype %in% 'ABC', ],
                     'tox_desc', 'tox_grade', n = n[1], zeros = '-'),
           tabler_by(worst[worst$subtype %in% 'GCB', ],
                     'tox_desc', 'tox_grade', n = n[2], zeros = '-'))
colnames(tt[[1]]) <- sprintf('Total<br /><font size=1>n = %s</font>', sum(n))
out <- do.call('cbind', tt)
out <- out[order(as.numeric(out[, 1]), decreasing = TRUE), ]

## no percent signs
out <- gsub('%', '', out)
```

Toxicities attributed to study treatement (possible, probable, definite) are presented in Table 3. All patients received at least on dose of study treatment and were included in toxicity summary. The most common toxicities on study were `r print_counts(rep(tolower(names(nn <- tail(sort(table(worst$tox_desc)), 10))), nn), sep = ';')`.

The highest overall grades experienced among all patients regardless of toxicity were `r print_counts(nn)`.

```{r, results='asis'}
htmlTable(out, cgroup = c('', tbl_headers[-1]), n.cgroup = sapply(tt, ncol),
  tfoot = paste0('<font size=1><sup>&dagger;</sup>Percentages represent ',
            'proportion of patients out of respective phase total.</font>'),
  caption = 'Table 3: Toxicities<sup>&dagger;</sup> by subtype, grade.')
```

## Outcome summary

```{r outcome}
## response
responders <- grepl('R', resp$resp)
pr_better <- resp$resp %in% c('PR','CR')
crs <- resp$resp %in% 'CR'


## survival models and plots
os_fit <- survfit(Surv(os_time, os_ind) ~ subtype,
                  data = dd, conf.type = 'log-log')
pfs_fit <- survfit(Surv(pfs_time, pfs_ind) ~ subtype,
                   data = dd, conf.type = 'log-log')
```

### Response to treatment

The primary outcome was overall response rate (MR, PR, CR) at end of treatment. All patients received at least one dose of study treatment and were included in response assessment, and all patients were evaluated for EOT response.

The overall response rate of MR or better was `r binconr(sum(responders), N)`. `r nnr(sum(pr_better))` patients experienced a PR or better at EOT (`r binconr(sum(pr_better), N, est = FALSE)`). `r nnr(sum(crs))` patients experienced a CR or better (`r binconr(sum(crs), N, est = FALSE)`).

### Time-to-event

Patients were followed for progression-free and overall survival, and results are reporte in Tables 4 and 5.

```{r survtbl, results='asis'}
pfs <- surv_table(pfs_fit, digits = 2)
os <- surv_table(os_fit, digits = 2)

f <- function(x, cap)
  htmlTable(do.call('rbind', x), caption = cap,
          tspanner = c('ABC','GCB'), n.tspanner = sapply(x, nrow))

f(pfs, cap = 'Table 4: Progression-free survival.')

f(os, cap = 'Table 5: Overall survival.')
```

```{r survfig, fig.width=8, fig.height=9, fig.cap='Time-to-event endpoints, profression-free (panel A) and overall (B) survival.'}
## survival plot wrapper
surv_plot <- function(x, ...) {
  rawr::kmplot(x, col.band = NA, col.surv = seq_along(x$strata),
               legend = FALSE, xlab = 'Time (months)',
               family = 'HersheySerif', ..., mar = c(6, 10, 2, 2))
}

mt <- function(x)
  mtext(side = 3, text = x, at = -5,
      family = 'HersheySerif', font = 2, cex = 1.5)
  

par(mfrow = c(2, 1), bty = 'l')
surv_plot(pfs_fit, ylab = 'PFS probability', add = TRUE)
mt('A')
surv_plot(os_fit, ylab = 'OS probability', add = TRUE)
mt('B')
box('outer')
```

## Discussion

Good programming practices and using `knitr` to create dynamic documents can make your life so much easier.

---

```{r session}
sessionInfo()
```

## References

Loprinzi CL. Laurie JA. Wieand HS. Krook JE. Novotny PJ. Kugler JW. Bartel J. Law M. Bateman M. Klatt NE. et al. "Prospective evaluation of prognostic variables from patient-completed questionnaires." North Central Cancer Treatment Group. *Journal of Clinical Oncology*. 12(3):601-7, 1994.

```{r, results='asis'}
## r citation
print(citation(), style = 'html')
print(citation(package = 'survival'), style = 'html')
```

## Appendix

```{r append, child='./99-append.Rmd'}
```
